Bird beak's elaboration and exploration

### 2021-09-17 Analysis/03-elaboration_exploration_analyses.Rmd
typo
Update 03-elaboration_exploration_analyses.Rmd

>TG: TODO: do this plot on a phylogeny

@@

With R_{\vec{B}} being the rotation matrix of the vector $\vec{B}$ onto $\vec{\hat{\imath}}$:

>>>>

```{r, echo = FALSE, eval = TRUE}
library(phytools)
source("~/Packaging/phytools/R/plotTree.wBars.R")

get.disparity.matrix <- function(data, col) {
    ## Get the matrices
    matrices <- lapply(data$disparity, `[[`, 1)
    ## Add the rownames
    list_of_names <- list()
    for(one_group in 1:length(matrices)) {
        rownames(matrices[[one_group]]) <- list_of_names[[one_group]] <- rownames(data$matrix[[1]])[c(data$subsets[[one_group]]$elements)]
    }
    ## Name the groups
    names(list_of_names) <- names(data$subsets)

    ## Matrix
    matrix <- do.call(rbind, matrices)

    ## Colours
    if(!missing(col)) {
        name.cols <- function(names, col) {
            out <- rep(col, length(names))
            names(out) <- names
            return(out)
        }
        col <- mapply(name.cols, list_of_names, as.list(col))
        names(col) <- NULL
        col <- unlist(col)
    }

    ## Output
    if(missing(col)) {
        return(list(data = matrix, names = list_of_names))
    } else {
        return(list(data = matrix, names = list_of_names, col = col))
    }
}

## Get the different elements
phylo_elaboration  <- get.disparity.matrix(results_phylo$position,
                                          col = colour_vector[-4])
within_elaboration <- get.disparity.matrix(
    get.subsets(results_within$position, subsets = c(1,2,3)),
                                          col = colour_vector[-4])
phylo_exploration  <- get.disparity.matrix(results_phylo$distance,
                                          col = colour_vector[-4])
within_exploration <- get.disparity.matrix(
    get.subsets(results_within$distance, subsets = c(1,2,3)),
                                          col = colour_vector[-4])
phylo_angle        <- get.disparity.matrix(results_phylo$degree,
                                          col = colour_vector[-4])
within_angle       <- get.disparity.matrix(
    get.subsets(results_within$degree, subsets = c(1,2,3)),
                                          col = colour_vector[-4])
```

## Projections (on the main phylogeny)

```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
## Plot per species
plotTree.wBars_modif(demo_tree, phylo_elaboration$data, type = "phylogram", col = phylo_elaboration$col, tip.label = TRUE, main = "Elaboration", mar = c(3, 0.1, 0.1, 0.1), args.distribution = list(show.scale = TRUE))
```

```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
## Plot per species
plotTree.wBars_modif(demo_tree, phylo_exploration$data, type = "phylogram", col = phylo_elaboration$col, tip.label = TRUE, main = "Exploration", mar = c(3, 0.1, 0.1, 0.1), args.distribution = list(show.scale = TRUE))
```

```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
## Plot per species
plotTree.wBars_modif(demo_tree, phylo_angle$data, type = "phylogram", col = phylo_elaboration$col, tip.label = TRUE, main = "Angle", mar = c(3, 0.1, 0.1, 0.1), args.distribution = list(show.scale = TRUE))
```

## Projections (onto each clade's phylogeny)

```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
## Plot per species
plotTree.wBars_modif(demo_tree, within_elaboration$data, type = "phylogram", col = phylo_elaboration$col, tip.label = TRUE, main = "Elaboration (within)", mar = c(3, 0.1, 0.1, 0.1), args.distribution = list(show.scale = TRUE))
```

```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
## Plot per species
plotTree.wBars_modif(demo_tree, within_exploration$data, type = "phylogram", col = phylo_elaboration$col, tip.label = TRUE, main = "Exploration (within)", mar = c(3, 0.1, 0.1, 0.1), args.distribution = list(show.scale = TRUE))
```

```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
## Plot per species
plotTree.wBars_modif(demo_tree, within_angle$data, type = "phylogram", col = phylo_elaboration$col, tip.label = TRUE, main = "Angle (within)", mar = c(3, 0.1, 0.1, 0.1), args.distribution = list(show.scale = TRUE))
```

These graphs can also be plotted as circular for the big trees later.

```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
## Plot per species
plotTree.wBars_modif(demo_tree, within_angle$data, type = "fan", col = phylo_elaboration$col, tip.label = FALSE, main = "Angle (within)", mar = c(3, 0.1, 0.1, 0.1), args.distribution = list(show.grid.scale = TRUE))
```

## Color PCA according to mean proj/rej?

And there's also this way of looking at it. But bof. Kinda ugly and probably too much info. (also it seems something is not working correctly).

```{r,echo = FALSE, eval = TRUE}
## Plotting fun
plot.gradient.space <- function(data, values, cent.tend = median, col, gradient.fun = heat.colors, ...) {

    get.colour.gradient <- function(values, col.fun, col.range, get.range = FALSE) {

        ## Sort the data by range
        histo <- hist(c(col.range, values), plot = FALSE)
        n_col <- length(histo$counts)

        ## Get the gradient
        avail_cols <- rev(col.fun(n_col))

        ## Attribute the colours
        col_out <- rep(avail_cols[n_col], length(values))
        range <- avail_cols[c(n_col, 1)]

        ## Attribute the colour by range
        for(colour in rev(1:n_col)) {
          col_out[(values <= histo$breaks[colour])] <- avail_cols[colour]
        }

        if(get.range) {
            return(list(range = range, col = col_out))
        } else {
            return(col_out)
        }
    }

    ## Get the colour gradient
    gradient_values <- apply(values, 1, cent.tend)
    gradient_cols <- get.colour.gradient(gradient_values, col.fun = gradient.fun, col.range = range(gradient_values), get.range = TRUE)
    colour_gradient <- gradient_cols$col
    names(colour_gradient) <- rownames(values)

    ## Plot the points
    sauron.plot(data,
                col = col,
                major.axes = mean,
                ellipses = mean,
                legend = TRUE,
                points = TRUE,
                pch = 19,
                cex = 1.5
                , ...)
    points(data$matrix[[1]], pch = 19, col = colour_gradient)
    legend("bottomright", pch = 19, col = gradient_cols$range, legend = round(range(gradient_values), 2))
}
```

```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
plot.gradient.space(data = covar_matrices, values = phylo_elaboration$data, col = colour_vector, main = "Elaboration gradient")
```
```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
plot.gradient.space(data = covar_matrices, values = phylo_exploration$data, col = colour_vector, main = "Exploration gradient")
```
```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
plot.gradient.space(data = covar_matrices, values = phylo_angle$data, col = colour_vector, main = "Angle gradient")
```

```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
plot.gradient.space(data = covar_matrices, values = within_elaboration$data, col = colour_vector, main = "Elaboration gradient (within)")
```
```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
plot.gradient.space(data = covar_matrices, values = within_exploration$data, col = colour_vector, main = "Exploration gradient (within)")
```
```{r, fig.height = 8, fig.width = 8, echo = FALSE, eval = TRUE}
plot.gradient.space(data = covar_matrices, values = within_angle$data, col = colour_vector, main = "Angle gradient (within)")
```

@@

With $R_{\vec{B}}$ being the rotation matrix of the vector $\vec{B}$ onto $\vec{\hat{\imath}}$:
